---
title: "Untitled"
format: html
editor: visual
---

# Practice Simple Linear Regression

First I am going to install the packages.

```{r}
install.packages(c("tidyverse", "skimr", "broom", "infer"))
```

Loading them,

```{r}
library(tidyverse)
library(skimr)
library(broom)
library(infer)
```

### Step 01

First we have to load the dataset and do a quick exploratory data analysis for the quantitative variables present in the dataset. First reading the dataset, and storing it ij a tibble called "d".

```{r}
f <- "https://raw.githubusercontent.com/difiore/ada-datasets/main/Street_et_al_2017.csv"

d <- read_csv(f)
```

First I have to identify what are the quantitative variables are present in the dataset. Checking the data we have,

```{r}
glimpse(d)
```

From what I can see there are two categorical variables and the rest are numeric. To select the data frame which only contains the quantitative variables,

```{r}
d_quantitative <-  d %>%
  select(where(is.numeric))

d_quantitative
```

As suggested, to get a quick summary, we can use skim(),

```{r}
skim(d_quantitative)
```

### Step 02

Now we need to plot brain size (ECV) as a function of social group size (Group_size), longevity (Longevity), juvenile period length (Weaning), and reproductive lifespan (Repro_lifespan).

I need four plots for this.

#### First, Let's plot ECV vs Group_size

```{r}
ggplot(d, aes(x = Group_size, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "ECV as a function of Group Size",
    x = "Group Size",
    y = "ECV"
  )
```

Next, I am going to plot ECV vs Longevity

```{r}
ggplot(d, aes(x = Longevity, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "ECV as a function of Longevity",
    x = "Longevity",
    y = "ECV"
  )
```

#### Then, we can plot ECV vs Weaning

```{r}
ggplot(d, aes(x = Weaning, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "ECV as a function of Weaning",
    x = "Weaning",
    y = "ECV"
  )
```

#### Finally, I have to plot ECV vs Reproductive lifespan

```{r}
ggplot(d, aes(x = Repro_lifespan, y = ECV)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "ECV as a function of Reproductive Lifespan",
    x = "Reproductive Lifespan",
    y = "ECV"
  )
```

### Step 03

Calculating ordinary least squares regression coefficients beta 1 and beta 0 for ECV as a function of social group size.

That means we need the regression where,

ECV = beta0 + beta1(Group_size)

As instructed I need to remove the rows from my data set where one of these variables is missing. I am using the initial data set for this.

```{r}

d1 <- d %>%
  select(ECV, Group_size) %>%
  drop_na()

d1
nrow(d1)
```

Calculating this by hand, from the formulas of slope = cov(x,y)/var(x) and Intercept (beta0) = y' - beta1(x')

```{r}
x <- d1$Group_size
y <- d1$ECV

beta1 <- cov(x, y) / var(x)
beta0 <- mean(y) - beta1 * mean(x)

beta1
beta0
```
